---
title: "Simulated Lateness Distributions"
author: "Clare Clingain"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

From the MTA data, I'll obtain lateness distributions which will show by how many minutes people miss the connection between the 1 train and the Staten Island ferry. Before looking at the empirical distributions, I'll first simulate what these distributions could look like given a set of assumptions. 

We first assume that the 1 train and the Staten Island ferry have fixed schedules. We can then work with two possible assumptions:

1) normal distribution of lateness with some variance (fixed through the day)

2) beta distribution of lateness. Since the beta distribution is positive, this means that trains will always be late (gets worse throughout the day)

# Data Structure

The simulated data will start off with two columns: time stamps for schedule and day number (1:n). The noise will be added as a third column, with the fourth column calculated as the final arrival times. 

# Simulations

```{r set up}
start_time <- as.POSIXct("2018-01-01 01:04:00")
# train is every 20 minutes until 5:40ish
diff <- 60*20
times <- NULL
for(i in 1:15){
  if(i == 1){
    times[i] <- start_time
  } else {
    prev <- i - 1
  times[i] <- times[prev] + diff
  }
}
# train is every 10 minutes until 7:25ish
diff <- 60*10
for(i in 16:27){
  prev <- i - 1
  times[i] <- times[prev] + diff
}
# train is every 4-6 minutes until 12:25ish (will go with 5 mins)
diff <- 60*5
for(i in 28:88){
  prev <- i - 1
  times[i] <- times[prev] + diff
}
# train is every 6 minutes until 3:51ish
diff <- 60*6
for(i in 89:119){
  prev <- i - 1
  times[i] <- times[prev] + diff
}
# train is every 4-6 minutes until 9:37ish (pm) (will go with 5 mins)
diff <- 60*5
for(i in 119:187){
  prev <- i - 1
  times[i] <- times[prev] + diff
}
# train is every 6 minutes until 11ish
diff <- 60*6
for(i in 187:200){
  prev <- i - 1
  times[i] <- times[prev] + diff
}
# train is every 10 minutes till 12:50ish
diff <- 60*10
for(i in 201:206){
  prev <- i - 1
  times[i] <- times[prev] + diff
}
# convert back to POSIX
class(times) <- c('POSIXt', 'POSIXct')
```

This will serve as our fixed 1 train schedule.
