---
title: "Wait Time Analysis"
author: "Clare Clingain"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(lubridate)
require(dplyr)
library(tidyverse)
load("../Data/Clean/analysis_data.RData")
```

# Research Questions

Here are the research questions I hope to answer:

1) What does the distribution of lateness look like for connections between the 1 train and the Staten Island Ferry?

2) How does this distribution vary across time (e.g., rush hour vs tourist peak)?

3) How does this distribution vary for people who can run versus people who can't run from the 1 train to the ferry?

# Initial Visualizations

## Density plots

```{r density}
# overlay plot
plot(density(analysis_data$wait_times_1min), main = "Wait times with 1, 2, and 3 minutes exit time", col = 2)
lines(density(analysis_data$wait_times_2min), col = 4)
lines(density(analysis_data$wait_times_3min), col = 5)
legend("topright",legend = c("1 min","2 min","3 min"), lty = 1, lwd = 2, col = c(2,4,5))
```

The three distributions look quite similar. In terms of the simulations I ran, these distributions look like the beta distribution results. 

## Time (hour) x Wait times

```{r time by wait times}
plot(as.factor(analysis_data$hour), analysis_data$wait_times_1min, main = "Wait times  (1 min exit time)",xlab = "Hour of day",ylab = "Wait time")
plot(as.factor(analysis_data$hour), analysis_data$wait_times_2min, main = "Wait times  (2 min exit time)",xlab = "Hour of day",ylab = "Wait time")
plot(as.factor(analysis_data$hour), analysis_data$wait_times_3min, main = "Wait times (3 min exit time)",xlab = "Hour of day",ylab = "Wait time")
```

It makes sense that the waits pile up late at night and during the tourist peak hours (10-3). I'm curious seeing that the very lengthy wait times caused by a missing ferry are so dispersed. I initially thought that it would be worse in the wee hours of the morning, but it looks to be a problem throughout the day. This isn't concerning in terms of something being wrong with the data because there are many reasons why a ferry or a train never shows. Both can break down, both can have sick passengers, and both can barely miss each other. 

```{r med iqr}
# medians
medians_IQR <- analysis_data %>% 
  group_by(as.factor(hour)) %>% 
  summarise(median_1min = median(wait_times_1min),
            median_2min = median(wait_times_2min),
            median_3min = median(wait_times_3min),
            IQR_1min = IQR(wait_times_1min),
            IQR_2min = IQR(wait_times_2min),
            IQR_3min = IQR(wait_times_3min))
colnames(medians_IQR)[1] <- "hour"
# plot the medians and IQR
plot(x = as.numeric(medians_IQR$hour), y = medians_IQR$median_1min, pch = 16, main = "Median wait times by hour", ylab = "Wait time (in minutes)", xlab = "Hour")
lines(x = as.numeric(medians_IQR$hour), y = medians_IQR$median_1min, lwd = 2)
lines(x = as.numeric(medians_IQR$hour), y = medians_IQR$median_2min, lwd = 2, col = 2, type = 'o', pch = 16)
lines(x = as.numeric(medians_IQR$hour), y = medians_IQR$median_3min, lwd = 2, col = 4, type = 'o', pch = 16)
legend("bottomright",legend = c("1 min","2 min","3 min"), lty = 1, lwd = 2, col = c(1,2,4))
```

For the rush-hour and tourist peak hours, it looks like time-to-exit doesn't make a difference. A longer time-to-exit does seem to matter for the late night/early morning hours. 

## Time (weekday) x Wait times

Now I'll check the wait times by weekday. 

```{r weekday by wait}
plot(analysis_data$weekday, analysis_data$wait_times_1min, main = "Wait times (1 min exit time)",xlab = "Day of the week",ylab = "Wait time")
plot(analysis_data$weekday, analysis_data$wait_times_2min, main = "Wait times (2 min exit time)",xlab = "Day of the week",ylab = "Wait time")
plot(analysis_data$weekday, analysis_data$wait_times_3min, main = "Wait times (3 min exit time)",xlab = "Day of the week",ylab = "Wait time")
```

I'm curious as to why the distributions for Tuesday and Thursday seem more spread and have slightly higher medians than Monday, Wednesday, and Friday. Once again, we see that the long delays seem to be evenly distributed across days of the week. I'd like to test if these distrbutions are different from each other later on. 

## Time (Month) x Wait times

```{r month wait time}
plot(as.factor(analysis_data$month), analysis_data$wait_times_1min, main = "Wait times (1 min exit time)",xlab = "Month",ylab = "Wait time")
plot(as.factor(analysis_data$month), analysis_data$wait_times_2min, main = "Wait times (2 min exit time)",xlab = "Month",ylab = "Wait time")
plot(as.factor(analysis_data$month), analysis_data$wait_times_3min, main = "Wait times (3 min exit time)",xlab = "Month",ylab = "Wait time")
```

## Time (Rush vs nonrush) x Wait times

```{r rush wait}
plot(analysis_data$rushour, analysis_data$wait_times_1min, main = "Wait times (1 min exit time)",xlab = "Day of the week",ylab = "Wait time")
plot(analysis_data$rushour, analysis_data$wait_times_2min, main = "Wait times (2 min exit time)",xlab = "Day of the week",ylab = "Wait time")
plot(analysis_data$rushour, analysis_data$wait_times_3min, main = "Wait times (3 min exit time)",xlab = "Day of the week",ylab = "Wait time")
```



# Test the difference between the wait distributions

I will use the two-sample Kolmogorov-Smirnov test to determine whether each distribution is different from the other two. I will adjust my p-values to reflect the multiple tests. 

```{r kolmogorov smirnov}
one_two_res <- ks.test(analysis_data$wait_times_1min, analysis_data$wait_times_2min)
one_two_res
one_three_res <- ks.test(analysis_data$wait_times_1min, analysis_data$wait_times_3min)
one_three_res
two_three_res <- ks.test(analysis_data$wait_times_2min, analysis_data$wait_times_3min)
two_three_res

# adjust p-values
paste("The adjusted p-values for each respective test are",p.adjust(c(one_two_res$p.value, one_three_res$p.value, two_three_res$p.value), method = "hochberg"))
```

We see a significant difference between all three distributions. Although this is great news, I strongly wonder if the data violate the iid assumption of the KS test. Even though some wait times may be independent of one another (i.e., morning rush from evening rush, January from June), I doubt that this is the case across all the data. Two consectutive trains are unlikely to be independent. Nevertheless, this test gives us an additional piece of information that we will use with caution. 

# Test difference between distribution means

As discussed above, the iid assumption is trick with this data. However, I'd like to get some 95% confidence intervals for the difference between the means.

```{r t tests}
t1 <- t.test(analysis_data$wait_times_1min, analysis_data$wait_times_2min)
t2 <- t.test(analysis_data$wait_times_1min, analysis_data$wait_times_3min)
t3 <- t.test(analysis_data$wait_times_2min, analysis_data$wait_times_3min)
paste("The 95% confidence interval for the difference in means between 1min vs 2min is (", round(t1$conf.int[1],3),",", round(t1$conf.int[2],3),")")
paste("The 95% confidence interval for the difference in means between 1min vs 3min is (", round(t2$conf.int[1],3),",", round(t2$conf.int[2],3),")")
paste("The 95% confidence interval for the difference in means between 2min vs 3min is (", round(t3$conf.int[1],3),",", round(t3$conf.int[2],3),")")
```

